Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels

### 2023-10-09 Delaneyetal2023b.tex
Updates from Overleaf

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels, and compares it to subaerial ones.
In the absence of continuous measurements of subglacial and subaerial water velocity and channel morphology, numerical parameterizations of subglacial and subaerial hydraulics are utilized.
Parameterizations are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The parameterizations' lumped nature isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
Furthermore, outputs are compared to sediment discharge data from these two catchments.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport from glacierized catchments.

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data interpolated using a linear spline.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Variability of an output is represented by the  standard deviation of an output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one half of the timescale before and after a point time.

@@

Over time scales of weeks or longer, the glacier's sediment transport capacity responds to the ice thickness, controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge \cite<Figure~\ref{fig:multi_run}, ~ Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022}.
In a supply-limited state, often assumed in landscape evolution models, then glaciers' sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically responds to sediment size, water discharge, and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,tucker1997,wickert2019}.
The stronger link between sediment transport capacity and hydrological forcing in subaerial channels may mean that variations in sediment transport records can be more easily attributed to climatic conditions than in subglacial systems.

@@

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
  Regime & Fixed & Determined via D-W
  \makecell{Equation~\ref{eq:DW}} & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
  Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & 
\makecell{\(w_c\,\propto Q^\alpha\) \\ \(S=w_cH\) \\ \(D_h=H\)} & 
\(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}

>>>>

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
In the absence of continuous measurements of subglacial and subaerial water velocity and channel morphology, numerical parameterizations of subglacial and subaerial hydraulics are utilized.
Parameterizations are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The parameterizations' lumped nature isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
Furthermore, outputs are compared to sediment discharge data from these two catchments.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport from glacierized catchments.

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data were interpolated using a linear spline.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{2}$.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The variability of output is represented by the  standard deviation of that output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one-half of the timescale before and after a point time.

@@

Over time scales of weeks or longer, the glacier's sediment transport capacity responds to the ice thickness, controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022}.
In a supply-limited state, often assumed in landscape evolution models, then glaciers' sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically responds to sediment size, water discharge, and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,tucker1997,wickert2019}.
The stronger link between sediment transport capacity and hydrological forcing in subaerial channels may mean that variations in sediment transport records can be more easily attributed to climatic conditions than in subglacial systems.

@@

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
  Regime & Fixed & \makecell{Determined via D-W
  \\Equation~\ref{eq:DW}} & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
  Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & 
\makecell{\(w_c\,\propto Q^\alpha\) \\ \(S=w_cH\) \\ \(D_h=H\)} & 
\(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}
